---
title: "Homework 5"
author: "Brandon Chung, Jiaxin Zheng, Andreina Arias, and Stephanie Chiang"
date: "Fall 2025"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
---

In this homework assignment, you will explore, analyze and model a data set containing information on
approximately 12,000 commercially available wines. The variables are mostly related to the chemical properties of
the wine being sold. The response variable is the number of sample cases of wine that were purchased by wine
distribution companies after sampling a wine. These cases would be used to provide tasting samples to
restaurants and wine stores around the United States. The more sample cases purchased, the more likely is a
wine to be sold at a high end restaurant. A large wine manufacturer is studying the data in order to predict the
number of wine cases ordered based upon the wine characteristics. If the wine manufacturer can predict the
number of cases, then that manufacturer will be able to adjust their wine offering to maximize sales.
Your objective is to build a count regression model to predict the number of cases of wine that will be sold
given certain properties of the wine. HINT: Sometimes, the fact that a variable is missing is actually predictive of
the target. You can only use the variables given to you (or variables that you derive from the variables provided).
Below is a short description of the variables of interest in the data set:


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```


```{r import}
train_df <- read.csv("https://github.com/Chung-Brandon/621-Data-Mining/blob/main/wine-training-data.csv")

test_df <- read.csv("https://raw.githubusercontent.com/Chung-Brandon/621-Data-Mining/refs/heads/main/wine-evaluation-data.csv")
```

